Is essentiality of genes conserved in Enterobacteriaceae?

### 2016-10-17 documents/manuscript.tex
Wrote abstract and conclusion

\section*{Abstract}
%Fitting multiple figures into very tight manuscripts while keeping it pleasant to read is challenging. Therefore figures are often simply attached to the very end of a manuscript file. While easier for the authors, this practice is inconvenient for readers. This \LaTeX template shows how to generate a compiled PDF with figures embedded into the text. It provides several examples of how to embed figures or tables directly into the text thus giving you a range of options from which you should choose the one best suited for your manuscript. Check out Schlegel et al., (2016) as example of use \cite{Schlegel2016}.
%
% now start line numbers
%\linenumbers

% the * after section prevents numbering
\section{Introduction}
With the advent of sequencing technologies and genome annotation methods many genes have been identified. However, not all of these genes are of the same importance for the growth of an organism. So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}. These studies can give us new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising minimal genomes \cite{hutchison_global_1999, hutchison_design_2016, reus_blueprint_2016}. Different methods have been used for studying the essentility of genes in prokaryotes. Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}. Another method that is widely used due to its simplicity and accuracy is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}. In this method, pools of single insertion mutants are constructed using transposon mutagenesis and the effect of each mutation on the survival of mutants is evaluated by sequencing the survivors \cite{barquist_approaches_2013}. This can lead to the identification of essential genes.

@@

Our aim is to study the essentiality of genes in an evolutionary framework in 13 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}). Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study. In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential. We first performed a detailed study of biases than can influence the inference of essentiality. Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome). We have also investigated how essentiality changes in the phylogenetic tree for these organisms.

\begin{figure}
\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in Keio collection \cite{baba_construction_2006}. We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
\label{fig:species-tree}
\end{figure}

@@

To study which functions are enriched in each class of essentiality, we gathered the descriptions of genes from their embl files. Then, we counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values. The P-values are then corrected using Benjamini-Hochberg-Yekutieli procedure. Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class. The results show an enrichment of the genes related to replication, transcription, translation, division, and rod shape determining proteins in essential class. The non-essential genes are mostly membrane associated proteins, flagellar proteins, ATPase, and DNA repair proteins. Beneficial losses are enriched in transposase enzymes, putative and hypothetical proteins, and mobile elements. Beneficial losses also contain many fimbrial proteins which probably has occurred because these proteins are not needed in a rich lab medium \textcolor{red}{\{TRUE?\}}.
%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}. It shows that most of the beneficial losses are mobile elements. Most of these mobile elements are not included in KEGG database.

We also conducted a pathway enrichment analysis for these three groups. For this, we downloaded pathway datasets for strains that were available in KEGG database. This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2. Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class. Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli. The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes. However, as mobile genetic elements are not stored in KEGG database, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.

@@

In this section, we have counted the genes that are essential and conserved at each level of the phylogenetic tree. We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree. In other words, we have tested if organisms close together have more essential genes in common than organisms that have separated earlier in the phylogenetic tree.

@@

As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods. For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods. Then, we performed word enrichment analysis that was explained before on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method. Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure. The results are depicted in Fig.\@\ref{fig:method-difference}.

@@

\subsection{Conservation classes}\label{sec:conservationclasses}
To study whether each gene in the 13 organisms is conserved we proposed a program that clusters homologous proteins. This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences. It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer. Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences. In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters. Finally, the program merges overlapping sequences in each cluster and combines similar clusters. The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 13 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.

@@

@@

>>>>

\section*{Abstract}
Many genes have been identified with advances in sequencing technology and genome annotation methods. However, not all of these genes are of the same importance. We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae. We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes. After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level. Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.

% now start line numbers
%\linenumbers

% the * after section prevents numbering
\section{Introduction}
With the advent of sequencing technologies and genome annotation methods many genes have been identified. However, not all of these genes are of the same importance for the growth of an organism. So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}. These studies can give us new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising minimal genomes \cite{hutchison_global_1999, hutchison_design_2016, reus_blueprint_2016}. Different methods have been used for studying the essentility of genes in prokaryotes. Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}. Another method that is widely used due to its simplicity is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}. In this method, pools of single insertion mutants are constructed using transposon mutagenesis and the effect of each mutation on the survival of mutants is evaluated by sequencing the survivors \cite{barquist_approaches_2013}. This can lead to the identification of essential genes.

@@

Our aim is to study the essentiality of genes in an evolutionary framework in 14 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}). Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study. In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential. We first performed a detailed study of biases than can influence the inference of essentiality. Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome). We have also investigated how essentiality changes in the phylogenetic tree for these organisms.

\begin{figure}
\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in EcoGene \cite{zhou_ecogene_2013}. We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
\label{fig:species-tree}
\end{figure}

@@

To study which functions are enriched in each class of essentiality, we used the word enrichment method explained in Section~\ref{sec:enrichment}. Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class. The results show an enrichment of the genes related to replication, transcription, translation, division, and rod shape determining proteins in essential class. The non-essential genes are mostly membrane associated proteins, flagellar proteins, ATPase, and DNA repair proteins. Beneficial losses are enriched in transposase enzymes, putative and hypothetical proteins, and mobile elements. Beneficial losses also contain many fimbrial proteins which probably has occurred because these proteins are not needed in a rich lab medium \textcolor{red}{\{TRUE?\}}.
%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}. It shows that most of the beneficial losses are mobile elements. Most of these mobile elements are not included in KEGG database.

We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}. The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes. However, as mobile genetic elements are not stored in KEGG database, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.

@@

In this section, we compared the number of genes that were essential and conserved at each level of the phylogenetic tree. We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree. In other words, we have tested if organisms close together have more essential genes in common than organisms that have separated earlier in the phylogenetic tree.

@@

As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods. For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods. Then, we performed word enrichment analysis explained in Section~\ref{sec:enrichment} on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method. Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure. The results are depicted in Fig.\@\ref{fig:method-difference}.

@@

\subsection{Conservation classes}\label{sec:conservationclasses}
To study whether each gene in the 14 organisms is conserved we proposed a program that clusters homologous proteins. This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences. It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer. Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences. In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters. Finally, the program merges overlapping sequences in each cluster and combines similar clusters. The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 14 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.

@@

\subsection{Word and pathway enrichment analysis}\label{sec:enrichment}
To study the enrichment of words in a group of genes compared to a second group, We gathered the descriptions of genes from their embl files. Then, we counted the repeat number of each word for the genes in each group and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values. The P-values were then corrected using Benjamini-Hochberg-Yekutieli procedure.

For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database. This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2. Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class. Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.

@@

\section{Conclusion}
In this paper, we studied the relationship between the essentiality and conservation of genes in 14 bacteria from Enterobacteriaceae family. We first studied the biases that can affect our results and showed that transposon insertions are more abundant near the origin of replication. In addition, there is a slight preferred insertion motif bias and a G-C bias in A-T rich genes. Moreover we showed that transposon insertions are more abundant near the ends of essential genes compared to the internal region, while it is the opposite in beneficial losses.

After correcting biases, we studied the essentiality versus conservation by dividing the genes into three classes of conservation: genus specific, single copy, and multi-copy, dividing them into four levels of essentiality: essential, ambiguous, non-essential, and benefical losses. We found that essential genes are mostly single copy, however there is a considerable number of genus specific genes in essential level which help to distinguish between genera. Furthermore, beneficial losses are mostly genus specific which means they are mostly new genes. The other finding was that multi-copy genes are mostly non-essential which can happen due to redundancy.

We also found that the pattern of essentiality changes is not in agreement with the phylogenetic tree due to the small number of essential genes. Also the ratio between core essential genes and core genes can be either constant or increasing depending on the method that we use for defining core and core essential genes. \textcolor{red}{\{WHAT?\}}

Overall, we have found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.
